---
title: 'STATS415: Data Mining Labs'
author: "Marian L. Schmidt"
date: "January 15, 2016"
output: html_document
---


### Lab 1:  January 15th, 2016
Introduction to R

We are using Introduction to Statistical Learning by James, Witten, Hastie, and  Tibshirani.  Online resources .
Book for statistical computing: 
<a href="http://www-bcf.usc.edu/~gareth/ISL/data.html" target="_blank">can be found here</a>   

We will be using the ISLR packages which was published with the book: <a href="https://cran.r-project.org/web/packages/ISLR/ISLR.pdf" target="_blank">ISLR Package Vignette</a>   



```{r}
x <- c(1,2,3,4,5)
length(x)
x <- c(1,6,2)
length(x)

j <- matrix(data = c(1, 2, 3, 4), nrow = 2, ncol = 2)
j
r <- matrix(c(5, 6, 7, 8), 2, 2)
r

set.seed(2016)
x <- rnorm(100)
y <- rnorm(100)
mean(x)
mean(y)
plot(x,y)

# How to save a figure
#pdf("Figure.pdf")
plot(x,y,col="red")
#dev.off()


x=seq(-pi,pi,length=50)
y=x
f=outer(x,y,function(x,y)cos(y)/(1+x^2))
contour(x,y,f)
contour(x,y,f,nlevels=45,add=T)
fa=(f-t(f))/2
contour(x,y,fa,nlevels=15)
image(x,y,fa)

persp(x,y,fa)
persp(x,y,fa,theta=30) # theta angles the viewing direction
persp(x,y,fa,theta=30,phi=20) # phi = colatitu
persp(x,y,fa,theta=30,phi=70)
persp(x,y,fa,theta=30,phi=40)


```


### Indexing Data


```{r}
A <- matrix(1:16,4,4, byrow = FALSE)
A
A[2,3]
A[c(1,3),c(2,4)]
A[1:3,2:4]
A[1:2,]
A[,1:2]
A[1,]
A[-c(1,3),]
A[-c(1,3),-c(1,3,4)]
dim(A)
```



# Loading Data

```{r}
Auto <- read.table("Auto.data.txt")
fix(Auto) # this opens up a txt file to edit the data!  
Auto <- read.table("Auto.data.txt",header=T,na.strings="?")
fix(Auto)
dim(Auto)
Auto[1:4,]
Auto <- na.omit(Auto) # Delete rows that have NAs
dim(Auto)
names(Auto)
```


# Writing Data

```{r}
# write.table(Auto, file = "newauto.txt", col.names = TRUE, row.names = FALSE)
```



# Additional Graphical and Numerical Summaries

```{r}
plot(cylinders, mpg)
plot(Auto$cylinders, Auto$mpg)
#attach(Auto)
plot(cylinders, mpg)
cylinders <- as.factor(cylinders)
plot(cylinders, mpg)
plot(cylinders, mpg, col="red")
plot(cylinders, mpg, col="red", varwidth=T)
plot(cylinders, mpg, col="red", varwidth=T,horizontal=T)
plot(cylinders, mpg, col="red", varwidth=T, xlab="cylinders", ylab="MPG")
hist(mpg)
hist(mpg,col=2)
hist(mpg,col=2,breaks=15)
pairs(Auto)
pairs(~ mpg + displacement + horsepower + weight + acceleration, Auto)
plot(horsepower,mpg)
identify(horsepower,mpg,name)
summary(Auto)
summary(mpg)
```


```{r set ggplot theme}
library(ggplot2)
library(cowplot)

theme_set(theme_bw() + theme(plot.title = element_text(face="bold", size = 12),  #Set the plot title
                                 strip.text.x = element_text(size=10, face="bold"),  #Set Y facet titles 
                                 strip.text.y = element_text(size=10, face="bold"),  #Set X facet titles 
                                 strip.background = element_blank(),  #No facet background
                                 axis.title.x = element_text(face="bold", size=12),  #Set the x-axis title
                                 axis.title.y = element_text(face="bold", size=12),  #Set the y-axis title
                                 axis.text.x = element_text(colour = "black", size=8),  #Set the x-axis labels
                                 axis.text.y = element_text(colour = "black", size=8),  #Set the y-axis labels
                                 legend.title = element_text(size = 8, face="bold"),  #Set the legend title 
                                 legend.text = element_text(size = 8),  #Set the legend text
                                 legend.position="right", #Default the legend position to the right
                                 plot.margin = unit(c(0.1, 0.1, 0.1, 0.1), "cm"))) #top, right, bottom, left
```


```{r chapter 2}

advert <- read.csv("Advertising.csv",header=T)
str(advert)

p1 <- ggplot(advert, aes(x = TV, y = Sales)) + geom_point(color = "orangered") + ggtitle("TV")

p2 <- ggplot(advert, aes(x = Radio, y = Sales)) + geom_point(color = "cornflowerblue") + ggtitle("Radio")

p3 <- ggplot(advert, aes(x = Newspaper, y = Sales)) + geom_point(color = "purple") + ggtitle("Newspaper")


plot_grid(p1, p2, p3, labels = c("A", "B", "C"), nrow = 1)

```

