---
title: "STATS 415 - Homework 5"
author: "Marian L. Schmidt"
date: "February 24, 2016"
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
output: pdf_document
---


**1.  The textbook describes that the `cv.glm()`function can be used in order to compute the LOOCV test error estimate. Alternatively, one could compute those quantities using just the `glm()` and `predict.glm()` functions, and a for loop. You will now take this approach in order to compute the LOOCV error for a simple logistic regression model on the Weekly data set.**

```{r}
# Load ISLR Library
library(ISLR)

# Set the Seed for reproducibility
set.seed(232)

# Look at what the data structure looks like
# How many rows and columns?
nrow(Weekly) # Number of Rows
ncol(Weekly) # Number of columns

#What does the data look like?
head(Weekly)

# What type of data do we have?
str(Weekly)
```

**(a) Fit a logistic regression model that predicts Direction using Lag1 and Lag2.**
```{r}
# Fit the logistic regression with all the data 
logistic_all <-  glm(Direction ~ Lag1 + Lag2, data = Weekly, family=binomial)
summary(logistic_all)
```

**(b) Fit a logistic regression model that predicts Direction using Lag1 and Lag2 using all but the first observation.**

```{r}
# Fit the logistic regression with all the data except one observation 
logistic_one <-  glm(Direction ~ Lag1 + Lag2, data = Weekly[-1,], family = binomial)
summary(logistic_one)
```

**(c) Use the model from (b) to predict the direction of the first observation. You can do this by predicting that the first observation will go up if `Pr(Direction=“Up” | Lag1, Lag2) > 0.5`. Was this observation correctly classified?**

```{r}
# Create a prediction regarding the market
logistic_prediction_up <- predict.glm(logistic_one, Weekly[1,], type = "response") > 0.05

# Is the market predicted to go up for the first observation?
logistic_prediction_up

# Does the market actually go up?
true_up <- Weekly[1, ]$Direction == "Up"

# Does the prediction match the market reality?
logistic_prediction_up != true_up
```
*This observation was incorrectly classified.  The prediction for the first observation is "Up", however, the actual observation was "Down".*


**(d) Write a for loop from `i=1` to `i=n`, where `n` is the number of observations in the data set, that performs each of the following steps:**  
  
  1. **Fit a logistic regression model using all but the ith observation to predict Direction using Lag1 and Lag2.**  
  2. **Compute the posterior probability of the market moving up for the ith observation.**  
  3. **Use the posterior probability for the ith observation in order to predict whether or not the market moves up.**  
  4. **Determine whether or not an error was made in predicting the direction for the ith observation. If an error was made, then indicate this as a 1, and otherwise indicate it as a 0.**  


```{r}
# Number of iterations in for loop
n <- nrow(Weekly); n
# Create 0s for error that we will fill in with ones
prediction_error <- rep(0, n)
for (i in 1:n){
  # Step 1: Run a logistic regression leaving one observation point out
  logistic_regression <- glm(Direction ~ Lag1 + Lag2, data = Weekly[-i,], family = binomial)
  # Step 2: Create a prediction on the one observation not included in the logistic regression
  market_pred_up <- predict.glm(logistic_regression, Weekly[i,], type = "response") > 0.05 
  # Step 3: Pull out all
  market_true_up <- Weekly[i, ]$Direction == "Up"  
  # Step 4: If an error was made in our prediction, add a "1" to error vector
  if(market_pred_up != market_true_up) prediction_error[i] <- 1 
}
prediction_error
sum(prediction_error)
```

*In the above example of LOOCV, there are `r sum(prediction_error)` misclassified predictions out of `r n`.*  

**(e) Take the average of the `n` numbers obtained in (d)iv in order to obtain the LOOCV estimate for the test error. Comment on the results.**

```{r}
mean(prediction_error) 
mean(prediction_error) * 100
```

*The LOOCV estimate for the test error rate is `r mean(prediction_error) * 100`%*
